\tit

### 2021-02-19 paper.tex
related work

A simple characterization of languages (and indeed functions from lists to any type) captures the essential idea behind language ``differentiation'' and ``integration'' as used for recognizing languages and leads to a collection of lemmas about type-level predicates.
These lemmas form the backbone of two dual implementations of language recognition---(inductive) regular expressions and (coinductive) tries---each containing the same code but in dual arrangements.
The language lemmas form most of the implementation in both cases, while the representation and primitive operations trade places.
The regular expression version corresponds to symbolic differentiation, while the trie version corresponds to automatic differentiation.

@@

In order to convert languages into parsers, it will help to decompose languages---and indeed any function from lists---into a regular form.
Since lists are defined by an algebraic data type, we can decompose languages according to constructor, as follows:
\agda{νδ}
Each use of \AF{δ} takes us one step closer to reducing general language membership to nullability.
In other words,\footnote{Repeated application is expressed as a standard left fold\out{ \stdlibCitep{Data.List}}:
\agda{foldl}
}%
\agda{ν∘foldlδ}

This simple observation is the semantic heart of the syntactic technique of \emph{derivatives of regular expressions} as used for efficient recognition of regular languages \citep{Brzozowski64}, later extended to parsing general context-free languages \citep{Might2010YaccID}.
Lemma \AF{ν∘foldlδ} liberates this technique from the assumption that languages are represented \emph{symbolically}, by some form of grammar (e.g., regular or context-free), inviting other representations as we will see in \secref{Automatic Differentiation}.

@@

The language-building vocabulary defined in \secref{Specifying Languages} can be simplified and generalized.
First note that there are two categories of operations.
One category transforms the codomain (types): \AF{∅}, \AF 𝒰, \AF{\_∪\_}, \AF{\_∩\_}.
The other transforms the domain (lists): \AF{𝟏}, \AF{\_⋆\_}, \AF{\_☆}, \AF{`}.
The first category applies to predicates over all types (not just over lists).
Within the second category, \AF{𝟏}, \AF{\_⋆\_}, and \AF{\_☆} apply to all monoids, while \AF{`} is specific to lists.

@@

\item \citet{AgularMannaa2009} also defined a non-indexed algebraic type of regular expressions.
They also \emph{defined} indexed types for \AF{ν} and \AF{δ} on regular expressions rather than defining the meanings of those operations in terms of languages and then proving properties of them.
They do prove a consistency property about \AF{δ}, but apparently not about \AF{ν}, and they note their method's ``inability to prove the non-membership of some string in a given language''.

>>>>

A simple characterization of languages (and indeed functions from lists to any type) captures the essential idea behind language ``differentiation'' (and integration) as used for recognizing languages and leads to a collection of lemmas about type-level predicates.
These lemmas form the backbone of two dual implementations of language recognition---(inductive) regular expressions and (coinductive) tries---each containing the same code but in dual arrangements.
The language lemmas form most of the implementation in both cases, while the representation and primitive operations trade places.
The regular expression version corresponds to symbolic differentiation, while the trie version corresponds to automatic differentiation.

@@

In order to convert languages into parsers, it will help to decompose languages---and indeed any function from lists---into a regular form.
Since lists are defined by an algebraic data type, we can decompose languages according to constructor, as follows:
\agda{νδ}
In terms of languages, {\AF{δ} \AB{P} \AB{a}} comprises all strings \AB{w} such that {\AB{a} \AIC ∷ \AB{w}} is in \AB{P}, i.e., the tails of strings in \AB{P} that start with \AB{a}.

Each use of \AF{δ} takes us one step closer to reducing general language membership to nullability.
In other words,\footnote{Repeated application is expressed as a standard left fold\out{ \stdlibCitep{Data.List}}:
\agda{foldl}
}%
\agda{ν∘foldlδ}
This simple observation is the semantic heart of the syntactic technique of \emph{derivatives of regular expressions} as used for efficient recognition of regular languages \citep{Brzozowski64}, later extended to parsing general context-free languages \citep{Might2010YaccID}.
Lemma \AF{ν∘foldlδ} liberates this technique from the assumption that languages are represented \emph{symbolically}, by some form of grammar (e.g., regular or context-free), inviting other representations as we will see in \secref{Automatic Differentiation}.

@@

The language-building vocabulary defined in \secref{Specifying Languages} can be simplified and generalized.
First note that there are two categories of operations.
One category transforms the codomain (types): \AF{∅}, \AF 𝒰, \AF{\_∪\_}, and \AF{\_∩\_}.
The other transforms the domain (lists): \AF{𝟏}, \AF{\_⋆\_}, \AF{\_☆}, and \AF{`}.
The first category applies to predicates over all types (not just over lists).
Within the second category, \AF{𝟏}, \AF{\_⋆\_}, and \AF{\_☆} apply to all monoids, while \AF{`} is specific to lists.

@@

\item \citet{AgularMannaa2009} also defined a non-indexed algebraic type of regular expressions.
They also \emph{defined} indexed types for \AF{ν} and \AF{δ} on regular expressions rather than defining the meanings of those operations in terms of languages and then proving properties of them.
They prove a consistency property about \AF{δ}, but apparently not about \AF{ν}, and they note their method's ``inability to prove the non-membership of some string in a given language''.
